Certified Knowledge Compilation \\ with Application to Formally Verified Model Counting

### 2023-08-20 doc/JAIR/jair-paper.tex
Incorporated example + optimizations into paper

\title{Certified Knowledge Compilation \\ with Application to Verified Model Counting}

@@

Generating proofs of unsatisfiability in SAT solvers has a long
tradition~\cite{ZhangMalik} and has become widely accepted due to the
formulation of clausal proof systems for which proofs can readily be
generated and efficiently checked
\cite{heule:cade:2013,wetzler14_drattrim}.
A number of formally verified checkers have been developed within different verification frameworks~\cite{cruz-cade-2017,lrat,Lammich:20,Tan:2021}.
The associated proofs add clauses while preserving satisfiability until the empty clause is derived.
Our work builds on the well-established technology and tools associated with clausal proof systems,
but we require features not found in proofs of unsatisfiability.
Our proofs construct a new propositional formula, and we must verify
%both
that it
%satisfies a set of rules and that it
is equivalent to the input formula.  This requires verifying
additional proof steps, including clause deletion steps, and subtle
invariants, as described in Sections~\ref{sect:cpog} and
\ref{section:formally-verified-toolchain}.

@@

Many important properties of Boolean formulas can be
expressed as ring evaluation problems.  The
(standard) \emph{model counting} problem for formula $\phi$ requires determining $|\modelset(\phi)|$.
It can be cast as a ring evaluation problem by having $\radd$ and
$\rmul$ be addition and multiplication over rational numbers and using
weight function $w(x) = 1/2$ for every variable $x$.
Ring evaluation of formula $\phi$ gives the \emph{density} of
the formula, i.e., the fraction of all possible total assignments that are
models.  For $n = |\varset|$, scaling the density by $2^n$
yields the number of models.  This formulation avoids the need for a ``smoothing'' operation,
in which redundant expressions are inserted into the formula~\cite{darwiche:jair:2002}.

@@

%We compare ring evaluation to algebraic model counting in Section~\ref{sect:future}.

@@

@@

The supplement~\cite{bryant:sat:2023:supplement},
shows the CPOG description for
an input formula with five clauses, yielding a POG with six
nonterminal nodes.  It explains how the clause addition and deletion
steps yield a proof of equivalence between the input formula and its POG
representation.

@@

For input formula $\inputformula$ and its translation into a POG $P$
with root node $\noder$, the most challenging part of the proof is to
show that $\modelset(\inputformula) \subseteq \modelset(\phi_\noder)$, i.e.,
that any total assignment $\assign$ that is a model of $\inputformula$
and the POG definition clauses
will yield $\assign(r) = 1$, for root literal $r$.  This part of the
proof consists of a series of clause assertions leading to one adding
$\{r\}$ as a unit clause.  We have devised two methods of generating this
proof.  The \emph{monolithic} approach makes just one call to a
proof-generating SAT solver and has it determine the relationship
between the two representations.  The \emph{structural} approach
recursively traverses the POG, generating proof obligations at each
node encountered.  It may require multiple calls to a proof-generating SAT
solver.

@@

@@

The performance of the structural proof generator for forward implication, both in its execution time and
the size of the proof generated, can be improved by two optimizations
described here.  A key feature is that they do not require any changes
to the proof framework---they build on the power of extended
resolution to enable new logical structures.  They
involve declaring new product nodes to encode products of literals.
These nodes are not part of the POG
representation of the formula; they serve only to enable the forward
implication proof.  Here we summarize the two optimizations.
The supplement~\cite{bryant:sat:2023:supplement} provides more details.

\emph{Literal Grouping:} A single recursive step of $\validate$ can encounter product nodes
having many---tens or even hundreds---of literals as children.  The
earlier formulation of $\validate$ considers each literal $\lit \in \lambda$
separately, calling a SAT solver for every literal that cannot be
validated with BCP\@.  Literal grouping handles
all of these literals together.  It defines product node $\nodev$
having the literals as children.  The goal then becomes to prove that
any total assignment must yield 1 for extension variable $v$.  Calling
a solver with $v$ set to 0 yields an unsatisfiability proof that can
be mapped back to a sequence of clause additions in the CPOG file validating
all of the literals.

\emph{Lemmas:} Our formulation of $\validate$ requires each call at a
node $\nodeu$ to recursively validate all of its children.  This
effectively expands the graph into a tree, potentially requiring an
exponential number of recursive steps.  Instead, for each node
$\nodeu$ having $\indegree(\nodeu) > 1$, the program can define and
generate the proof of a \emph{lemma} for $\nodeu$ when it is first reached by
a call to $\validate$ and then apply this lemma for this and
subsequent calls.  The lemma states that the POG with root node
$\nodeu$ satisfies forward implication for a formula
$\gamma_{\nodeu}$, where some of the clauses in this formula are
input clauses from $\inputformula$, but others are simplified
versions of input clauses.
The key idea is to introduce product nodes to encode (via DeMorgan's
Laws) the simplified clauses and have these serve as lemma arguments.

@@

>>>>

\title{Certified Knowledge Compilation \\ with Application to Formally Verified Model Counting}

@@

Generating proofs of unsatisfiability in SAT solvers has a long
tradition~\cite{ZhangMalik} and has become widely accepted due to the
formulation of clausal proof systems for which proofs can readily be
generated and efficiently checked
\cite{heule:cade:2013,wetzler14_drattrim}.
A number of formally verified checkers have been developed within different verification frameworks~\cite{cruz-cade-2017,lrat,Lammich:20,Tan:2021}.
The associated proofs add clauses while preserving satisfiability until the empty clause is derived.
Our work builds on the well-established technology and tools associated with clausal proof systems,
but we require features not found in proofs of unsatisfiability.
Our proofs construct a new propositional formula, and we must verify
both
that it
satisfies a set of rules and that it
is equivalent to the input formula.  This requires verifying
additional proof steps, including clause deletion steps, and subtle
invariants, as described in Sections~\ref{sect:cpog} and
\ref{section:formally-verified-toolchain}.

@@

Many important properties of Boolean formulas can be
expressed as ring evaluation problems.  The
(standard) \emph{model counting} problem for formula $\phi$ requires determining $|\modelset(\phi)|$.
It can be cast as a ring evaluation problem by having $\radd$ and
$\rmul$ be addition and multiplication over rational numbers and using
weight function $w(x) = 1/2$ for every variable $x$.
Ring evaluation of formula $\phi$ gives the \emph{density} of
the formula, i.e., the fraction of all possible total assignments that are
models.  For $n = |\varset|$, scaling the density by $2^n$
yields the number of models.

@@

@@

Kimmig's formulation of algebraic model
counting~\cite{kimmig:jal:2017} is more general of ours.  It allows
the algebraic structure to be a \emph{semiring}, i.e., there need not
be additive inverses.  For example, it can have the sum operation be
$\max$ and the product operation be addition over the real numbers plus $\infty$.  This more general form
requires that the representation generated by the knowledge compiler
obey a property known as \emph{smoothness}~\cite{darwiche:jair:2002}.
Within our formulation, a partitioned-operation formula is smooth when
all arguments to each sum operation have identical dependency sets.
That is, every sum operation $\bigvee_{1 \leq i \leq k} \phi_i$ has
$\dependencyset(\phi_i) = \dependencyset(\phi_j)$ for $1 \leq i < j \leq k$.
Smoothness can be ensured by adding redundant formulas to
artificially introduce variables.  For example, if subformula $\phi_i$
lacks having variable $x$ in its dependency set, it can be replaced by
$(x \por \obar{x}) \pand \phi_i$.  The consideration of smoothness is
orthogonal to the issues addressed by this paper.  Given a smoothed
dec-DNNF graph generated by a knowledge compiler, our toolchain will
convert this into a smooth POG and verify its equivalence to the input
formula.

@@

\subsection{CPOG Example}
\label{sect:cpog:example}

\begin{figure}
\begin{minipage}{0.58\textwidth}
(A)  Input Formula\\[1.2ex]
\begin{tabular}{lll}
\toprule
\makebox[5mm]{ID} & \makebox[15mm]{Clauses} & \\
\midrule
1 & \texttt{-1 3 -4} & \texttt{0} \\
2 & \texttt{-1 -3 4} & \texttt{0} \\
3 & \texttt{3 -4} & \texttt{0}\\
4 & \texttt{1 -3 4} & \texttt{0} \\
5 & \texttt{-1 -2} & \texttt{0} \\
\bottomrule
\end{tabular}
\\[1.8ex]
(C) POG Declaration\\[1.2ex]
\begin{tabular}{llll}
\toprule
\makebox[5mm]{ID} & \multicolumn{2}{l}{CPOG line} & Explanation \\
\midrule
6 & \texttt{p 5 -3 -4} & \texttt{0} & $p_5 = \obar{x}_3 \pand \obar{x}_4$ \\
9 & \texttt{p 6 3 4} & \texttt{0} & $p_6 = x_3 \pand x_4$ \\
12 & \texttt{s 7 5 6 7 10} & \texttt{0} & $s_7 = p_5 \por p_6$ \\
15 & \texttt{p 8 -1 7} & \texttt{0} & $p_8 = \obar{x}_1 \pand s_7$ \\
18 & \texttt{p 9 1 -2 7} & \texttt{0} & $p_9 = x_1 \pand \obar{x}_2 \pand s_7$ \\
22 & \texttt{s 10 8 9 16 19} & \texttt{0} & $s_{10} = p_8 \por p_9$ \\
 & \texttt{r 10} && Root $r = s_{10}$\\
\bottomrule
\end{tabular}
\end{minipage}
\begin{minipage}{0.35\textwidth}
(B) POG Representation \\
\input{dd/eg4}
\end{minipage}
%% Break
\\[2.5ex]
\begin{minipage}{0.42\textwidth}
(D) Defining Clauses\\[1.2ex]
\begin{tabular}{llll}
\toprule
\makebox[5mm]{ID} & \multicolumn{2}{l}{Clauses} & Explanation \\
\midrule
\texttt{6} & \texttt{5 3 4} & \texttt{0} & Define $p_5$ \\
\texttt{7} & \texttt{-5 -3} & \texttt{0} & \\
\texttt{8} & \texttt{-5 -4} & \texttt{0} & \\
\midrule
\texttt{9} & \texttt{6 -3 -4} & \texttt{0} & Define $p_6$ \\
\texttt{10} & \texttt{-6 3} & \texttt{0} & \\
\texttt{11} & \texttt{-6 4} & \texttt{0} & \\
\midrule
\texttt{12} & \texttt{-7 5 6} & \texttt{0} & Define $s_7$ \\
\texttt{13} & \texttt{7 -5} & \texttt{0} & \\
\texttt{14} & \texttt{7 -6} & \texttt{0} & \\
\midrule
\texttt{15} & \texttt{8 1 -7} & \texttt{0} & Define $p_8$ \\
\texttt{16} & \texttt{-8 -1} & \texttt{0} & \\
\texttt{17} & \texttt{-8 7} & \texttt{0} & \\
\midrule
\texttt{18} & \texttt{9 -1 2 -7} & \texttt{0} & Define $p_9$ \\
\texttt{19} & \texttt{-9 1} & \texttt{0} & \\
\texttt{20} & \texttt{-9 -2} & \texttt{0} & \\
\texttt{21} & \texttt{-9 7} & \texttt{0} & \\
\midrule
\texttt{22} & \texttt{-10 8 9} & \texttt{0} & Define $s_{10}$ \\
\texttt{23} & \texttt{10 -8} & \texttt{0} & \\
\texttt{24} & \texttt{10 -9} & \texttt{0} & \\
\bottomrule
\end{tabular}
\end{minipage}
\begin{minipage}{0.49\textwidth}
(E) CPOG Assertions\\[1.2ex]
\begin{tabular}{llllll}
\toprule
\makebox[5mm]{ID} & \multicolumn{2}{l}{Clause} & \multicolumn{2}{l}{Hint} & Explanation \\
\midrule
\texttt{25} & \texttt{a 5 1 3} & \texttt{0} & \texttt{3 6} & \texttt{0} & $\obar{x}_1 \land \obar{x}_3 \imply p_5$ \\
\texttt{26} & \texttt{a 6 1 -3} & \texttt{0} & \texttt{4 9} & \texttt{0} & $\obar{x}_1 \land x_3 \imply p_6$ \\
\texttt{27} & \texttt{a 3 7 1} & \texttt{0} & \texttt{13 25} & \texttt{0} & $\obar{x}_3 \land \obar{x}_1 \imply s_7$  \\
\texttt{28} & \texttt{a 7 1} & \texttt{0} & \texttt{27 14 26} & \texttt{0} & $\obar{x}_1 \imply s_7$  \\
\texttt{29} & \texttt{a 8 1} & \texttt{0} & \texttt{28 15} & \texttt{0} & $\obar{x}_1 \imply p_8$  \\
\texttt{30} & \texttt{a 5 -1 3} & \texttt{0} & \texttt{1 6} & \texttt{0} & $x_1 \land \obar{x}_3 \imply p_5$ \\
\texttt{31} & \texttt{a 6 -1 -3} & \texttt{0} & \texttt{2 9} & \texttt{0} & $x_1 \land x_3 \imply p_6$ \\
\texttt{32} & \texttt{a 3 7 -1} & \texttt{0} & \texttt{13 30} & \texttt{0} & $\obar{x}_3 \land x_1 \imply s_7$  \\
\texttt{33} & \texttt{a 7 -1} & \texttt{0} & \texttt{32 14 31} & \texttt{0} & $x_1 \imply s_7$  \\
\texttt{34} & \texttt{a 9 -1} & \texttt{0} & \texttt{5 33 18} & \texttt{0} & $x_1 \imply p_9$  \\
\texttt{35} & \texttt{a 1 10} & \texttt{0} & \texttt{23 29} & \texttt{0} & $\obar{x}_1 \imply s_{10}$  \\
\texttt{36} & \texttt{a 10} & \texttt{0} & \texttt{35 24 34} & \texttt{0} & $s_{10}$ \\
\bottomrule
\end{tabular}
\\[1.5ex]
(F) Input Clause Deletions\\[1.2ex]
\begin{tabular}{lllll}
  \toprule
 \multicolumn{3}{l}{CPOG line} & Explanation\\
\midrule
 \texttt{d 1} & \texttt{36 8 10 12 16 21 22} & \texttt{0} & Delete clause 1 \\
 \texttt{d 2} & \texttt{36 7 11 12 16 21 22} & \texttt{0} & Delete clause 2 \\
 \texttt{d 3} & \texttt{36 8 10 12 17 19 22} & \texttt{0} & Delete clause 3 \\
 \texttt{d 4} & \texttt{36 7 11 12 17 19 22} & \texttt{0} & Delete clause 4 \\
 \texttt{d 5} & \texttt{36 16 20 22} & \texttt{0} &  Delete clause 5 \\
\bottomrule
\end{tabular}
\end{minipage}
\caption{Example formula (A), its POG representation (B), and its CPOG proof (C), (E), and (F)}
\label{fig:eg4:all}
\end{figure}

Figure \ref{fig:eg4:all} illustrates an example formula and shows how
the CPOG file declares its POG representation.  The input formula (A)
consists of five clauses over variables $x_1$, $x_2$, $x_3$, and
$x_4$.  The generated POG (B) has six nonterminal nodes representing
four products and two sums.  We name these by the node
type (product $\nodep$ or sum $\nodes$), subscripted by the ID of the
extension variable.
% We denote the extension variable for nodes $\nodep_i$ and $\nodes_j$ as $p_i$ and $s_j$, respectively.
  The first part of the CPOG file (C) declares
these nodes using clause IDs that increment by three or four,
depending on whether the node has two children or three.  The last two
nonzero values in the sum declarations are the hint providing the
required mutual exclusion proof.

\subsection{Node Declarations}

We step through portions of the file to provide a better understanding of the CPOG proof framework.
Figure
\ref{fig:eg4:all}(D) shows the defining clauses that are implicitly
defined by the POG operation declarations.  These do not appear in the
CPOG file.  Referring back to the declarations of the sum nodes in
Figure \ref{fig:eg4:all}(C), we can see that the declaration of node
$\nodes_7$ had clause IDs 7 and 10 as the hint.  We can see in Figure
\ref{fig:eg4:all}(A) that these two clauses form a RUP proof for the clause
$\obar{p}_5 \lor \obar{p}_6$, showing that the two children of $\nodes_7$
have disjoint models.  Similarly, node $\nodes_{10}$ is declared as having
clause IDs 16 and 19 as the hint.  These form a RUP proof for the clause
$\obar{p}_8 \lor \obar{p}_9$, showing that the two children of
$\nodes_{10}$ have disjoint models.

Figure \ref{fig:eg4:all}(E) provides the sequence of assertions
leading to unit clause 36, consisting of the literal $s_{10}$.  This clause indicates that $\nodes_{10}$ is implied by the input clauses, i.e.,
any total assignment $\assign$
satisfying the input clauses must have $\assign(s_{10}) = 1$.
Working backward, we can see that
clause 29 indicates that variable $p_8$ will be implied by the input
clauses when $\assign(x_1) = 0$, while clause 34 indicates that node $p_9$ will
be implied by the input clauses when $\assign(x_1) = 1$.  These serve as the
hint for clause 36.

Figure \ref{fig:eg4:all}(F) shows the RUP proof steps required to
delete the input clauses.  Consider the first of these, deleting
input clause $\obar{x}_1 \lor x_3 \lor \obar{x}_4$.  The requirement is to show
that there is no total assignment $\assign$ that falsifies this clause but assigns $\assign(s_{10}) = 1$.
The proof proceeds by first assuming that the clause is false, requiring
$\assign(x_1) = 1$, $\assign(x_3) = 0$, and $\assign(x_4) = 1$.  The hint then consist of unit
clauses (e.g., clause 36 asserting that $\alpha(p_{10}) = 1$) or
clauses that cause unit propagation.  Hint clauses 8 and 10 force the
assignments $\assign(p_5) = \assign(p_6) = 0$.  These, plus hint clause 12 force
$\assign(s_7) = 0$.  This, plus hint clauses 16 and 21 force $\assign(p_8) = \assign(p_9) = 0$, leading,
via clause 22, to $\assign(s_{10}) = 0$.  But this contradicts clause 36,
completing the RUP proof.  The deletion hints for the other input
clauses follow similar patterns---they work from the bottom nodes of
the POG upward, showing that any total assignment that falsifies the clause
must assign $\assign(s_{10}) = 0$.

Deleting the asserted clauses is so simple that we do not show it.  It
involves simply deleting the clauses from clause number 35 down to
clause number 25, with each deletion using the same hint as were used
to add that clause.  In the end, therefore, only the defining clauses
for the POG nodes and the unit clause asserting $s_{10}$ remain,
completing a proof that the POG is logically equivalent to the input
formula.

@@

For input formula $\inputformula$ and its translation into a POG $P$
with root node $\noder$, the most challenging part of the proof is to
show that $\modelset(\inputformula) \subseteq \modelset(\phi_\noder)$, i.e.,
that any total assignment $\assign$ that is a model of $\inputformula$
and the POG definition clauses
will yield $\assign(r) = 1$, for root literal $r$.  This part of the
proof consists of a series of clause assertions leading to one adding
$\{r\}$ as a unit clause.  We have devised two methods for generating this
proof.  The \emph{monolithic} approach makes just one call to a
proof-generating SAT solver and has it determine the relationship
between the two representations.  The \emph{structural} approach
recursively traverses the POG, generating proof obligations at each
node encountered.  It may require multiple calls to a proof-generating SAT
solver.

@@

The forward implication proof of Figure~\ref{fig:eg4:all}(E) was
generated by the structural approach.  Working from step 36 backward,
we can see that steps 35 and 36 complete the call to
$\validate(\nodes_{10}, \emptyset, \phi_I)$.  This call used $x_1$ as
the splitting variable, first calling $\validate(\nodep_{8},
\{\obar{x}_1\}, \emptyset, \phi_I)$, which completed with step 29, and
$\validate(\nodep_{9}, \{x_1\}, \phi_I)$, which completed with step
34.  We see that each of these calls required separate traversals of
nodes $\nodes_7$, $\nodep_6$, and $\nodep_5$, with the former yielding
proof steps 25--27 and the latter yielding proof steps 30--32.  This
demonstrates how our simple formulation of $\validate$ effectively
expands the graph into a tree.  This shortcoming is avoided by the use of lemmas, as is described in Section~\ref{sect:lemma}.

@@

The performance of the structural proof generator for forward implication, both in its execution time and
the size of the proof generated, can be improved by two optimizations
described here.  A key feature is that they do not require any changes
to the proof framework---they build on the power of extended
resolution to enable new logical structures.  They
involve declaring new product nodes to encode products of literals.
These nodes are not part of the POG
representation of the formula; they serve only to enable the forward
implication proof.

@@

\subsection{Literal Grouping}
\label{sect:grouping}

A single recursive step of $\validate$ can encounter product nodes
having many literals as children.  The naive formulation of $\validate$
considers each literal $\lit \in \lambda$ separately.
Literal grouping allows all literals to be validated with a single call to a SAT solver.
It collects those literals
$\lit_1, \lit_2, \ldots, \lit_m$ that cannot be validated by BCP and defines a
product node $\nodev$ having these literals as children.  The goal
then becomes to prove that any total assignment must yield 1 for extension
variable $v$.  A single call to the solver can generate this proof by invoking it on the formula
  $\simplify{\psi}{\passign} \cup \simplify{\theta_{v}}{\{ \obar{v} \}}$, which should be unsatisfiable.
  The proof steps can be mapped back into clause addition steps in the CPOG file, incorporating the
  input clauses and the defining clauses for $\nodev$ into the hints.

\subsection{Lemmas}
\label{sect:lemma}

As we have noted, the recursive calling of $\validate$ starting at
root $\noder$ effectively expands the POG into a tree, and this can
lead to an exponential number of calls.
These shared subgraphs arise when the knowledge compiler employs \emph{clause caching}
to detect that the simplified set of
clauses arising from one partial assignment to the literals matches that
of a previous partial assignment~\cite{darwiche:aaai:2002}.
When this dec-DNNF node is translated into POG
node $\nodeu$, the proof generator can assume (and also check), that
there is a simplified set of clauses $\gamma_{\nodeu}$
for which the subgraph with root $\nodeu$ is its POG representation.

The proof generator can exploit the sharing of subgraphs
by constructing and proving a \emph{lemma} for each node
$\nodeu$ having $\indegree(\nodeu) > 1$.  This proof shows that any
total assignment $\assign$ that satisfies formula $\gamma_{\nodeu}$ and the defining clauses for the POG must yield
$\assign(u) = 1$.  This lemma is then invoked for every node having
$\nodeu$ as a child.
As a result, the generator will make recursive calls during a call to $\validate$ only once for each node in the POG\@.

The challenge for implementing this strategy is to find a way to
represent the clauses for the simplified formula $\gamma_{\nodeu}$ in the CPOG file.  Some may be
unaltered input clauses, and these can be used directly.  Others,
however will be clauses that do not appear in the input formula.  We
implement these by adding POG product nodes to the CPOG file to create
the appropriate clauses.  Consider an \emph{argument} clause
$C \in \gamma_{\nodeu}$ with $C = \lit_1 \lor \lit_2 \lor \cdots \lor \lit_k$.  If we
define a product node $\nodev$ with arguments
$\obar{\lit}_1, \obar{\lit}_2, \ldots, \obar{\lit}_k$, 
we will introduce a defining clause
$v \lor \lit_1 \lor \lit_2 \lor \cdots \lit_k$.  We call this a {\em
  synthetic} clause having $\obar{v}$ as the \emph{guard literal}.
That is, a partial assignment $\passign$ such that $\passign(v) = 0$ will {\em
  activate} the clause, causing it to represent argument clause $C$.  On the other
hand, a partial assignment with $\passign(v) = 1$ will
cause the clause to become a tautology and therefore have no effect.

Suppose for every clause $C_j \in \gamma_{\nodeu}$ that does not correspond to
an input clause, we generate a synthetic clause $C'_j$ with guard literal
$\obar{v}_j$, for $1 \leq j \leq m$.  Let $\gamma'_{\nodeu}$ be the formula where each clause $C_j$ is replaced by synthetic clause $C'_j$,
while input clauses in $\gamma_{\nodeu}$ are left unchanged.
Let $\lassign = \{ \obar{v}_1, \obar{v}_2, \ldots, \obar{v}_m \}$.
Invoking $\validate(\nodeu, \lassign, \gamma'_{\nodeu})$
 will then prove a lemma, given by the target clause
 $u \lor v_1 \lor v_2 \lor \cdots \lor v_m$,
 showing that any total assignment $\assign$ that activates the synthetic clauses will have $\assign(u) = 1$.

Later, when node $\nodeu$ is encountered by a call to $\validate(\nodeu, \passign, \psi)$, we invoke the lemma
by showing that each synthetic clause
$C_j$ matches some simplified clause in $\simplify{\psi}{\passign}$.  More precisely,
for $1 \leq j \leq m$,
we use clause addition to assert the clause
$\obar{v}_j \lor \bigvee_{\lit \in \passign} \obar{\lit}$,
showing that synthetic clause $C_j$ will be activated.
Combining the lemma with these activations provides a derivation of the target clause for the call to $\validate$.

Observe that the lemma structure can be hierarchical, since a shared
subgraph may contain nodes that are themselves roots of shared
subgraphs.  Nonetheless, the principles described allow the
definition, proof, and applications of a lemma for each shared node in
the graph.  For any node $\nodeu$, the first call to
$\validate(\nodeu, \passign, \psi)$ may require further recursion,
but any subsequent call can simply reuse the lemma proved by the first call.

\subsection{Lemma Example}
\label{app:lemma:eg}

\begin{figure}
(A) Additional nodes\\[1.0em]
\begin{tabular}{llll}
\toprule
\makebox[5mm]{ID} & \multicolumn{2}{l}{CPOG line} & Explanation \\
\midrule
25 & \texttt{p 11 -3 4} & \texttt{0} & $v_{11} = \obar{x}_3 \pand {x}_4$ \\
28 & \texttt{p 12  3 -4} & \texttt{0} & $v_{12} = {x}_3 \pand \obar{x}_4$ \\
\bottomrule
\end{tabular}
\\[1.0em]
(B) Implicit Clauses\\[1.2em]
\begin{tabular}{llll}
\toprule
\makebox[5mm]{ID} & \multicolumn{2}{l}{Clauses} & Explanation \\
\midrule
\texttt{25} & \texttt{11 3 -4} & \texttt{0} & Argument clause $\{x_3 ,\, \obar{x}_4\}$, activated by $\obar{v}_{11}$ \\
\texttt{26} & \texttt{-11 -3} & \texttt{0} & \\
\texttt{27} & \texttt{-11 4} & \texttt{0} & \\
\midrule
\texttt{28} & \texttt{12 -3 4} & \texttt{0} & Argument clause $\{\obar{x}_3,\,  {x}_4\}$, activated by $\obar{v}_{12}$ \\
\texttt{29} & \texttt{-12 3} & \texttt{0} & \\
\texttt{30} & \texttt{-12 -4} & \texttt{0} & \\
\bottomrule
\end{tabular}
\\[1.0em]
(C) CPOG Assertions\\[1.0em]
\begin{tabular}{llllll}
\toprule
\makebox[5mm]{ID} & \multicolumn{2}{l}{Clause} & \multicolumn{2}{l}{Hint} & Explanation \\
\midrule
\multicolumn{6}{l}{Lemma Proof} \\
\texttt{31} & \texttt{a 5 11 12 3} & \texttt{0} & \texttt{25 6} & \texttt{0} & $(\obar{v}_{11} \land \obar{v}_{12}) \land \obar{x}_3 \imply p_{5}$ \\
\texttt{32} & \texttt{a 6 11 12 -3} & \texttt{0} & \texttt{28 9} & \texttt{0} & $(\obar{v}_{11} \land \obar{v}_{12}) \land {x}_3 \imply p_{6}$ \\
\texttt{33} & \texttt{a 3 7 11 12} & \texttt{0} & \texttt{13 31} & \texttt{0} & $(\obar{v}_{11} \land \obar{v}_{12}) \land \obar{x}_3 \imply s_{7}$ \\
\texttt{34} & \texttt{a 7 11 12} & \texttt{0} & \texttt{33 14 32} & \texttt{0} & $(\obar{v}_{11} \land \obar{v}_{12}) \imply s_{7}$ \\
\midrule
\multicolumn{6}{l}{Lemma Application \#1} \\
\texttt{35} & \texttt{a -11 1} & \texttt{0} & \texttt{26 27 3} & \texttt{0} & $\obar{x}_1 \imply \obar{v}_{11}$ \\
\texttt{36} & \texttt{a -12 1} & \texttt{0} & \texttt{29 30 4} & \texttt{0} & $\obar{x}_1 \imply \obar{v}_{12}$ \\
\texttt{37} & \texttt{a 7 1} & \texttt{0} & \texttt{35 36 34} & \texttt{0} & $\obar{x}_1 \imply s_7$ \\
\midrule
\texttt{38} & \texttt{a 8 1} & \texttt{0} & \texttt{37 15} & \texttt{0} & $\obar{x}_1 \imply p_8$ \\
\midrule
\multicolumn{6}{l}{Lemma Application \#2} \\
\texttt{39} & \texttt{a -11 -1} & \texttt{0} & \texttt{26 27 1} & \texttt{0} & ${x}_1 \imply \obar{v}_{11}$ \\
\texttt{40} & \texttt{a -12 -1} & \texttt{0} & \texttt{29 30 2} & \texttt{0} & ${x}_1 \imply \obar{v}_{12}$ \\
\texttt{41} & \texttt{a 7 -1} & \texttt{0} & \texttt{39 40 34} & \texttt{0} & ${x}_1 \imply s_7$ \\
\midrule
\texttt{42} & \texttt{a 9 -1} & \texttt{0} & \texttt{5 41 18} & \texttt{0} & ${x}_1 \imply p_9$ \\
\texttt{43} & \texttt{a 1 10} & \texttt{0} & \texttt{23 38} & \texttt{0} & $\obar{x}_1 \imply s_{10}$  \\
\texttt{44} & \texttt{a 10} & \texttt{0} & \texttt{43 24 42} & \texttt{0} & $s_{10}$ \\
\bottomrule
\end{tabular}
\caption{Example of lemma definition, proof, and application}
\label{fig:eg4:lemmas}
\end{figure}

Figure~\ref{fig:eg4:lemmas} shows an alternate forward implication
proof for the example of Figure~\ref{fig:eg4:all} using a lemma to
represent the shared node $\nodes_7$.  We can see that the POG with
this node as root encodes the Boolean formula $x_3 \leftrightarrow x_4$, having a CNF representation consisting of the clauses
$\{x_3 ,\, \obar{x}_4\}$ and $\{\obar{x}_3 ,\, {x}_4\}$.  The product node
declarations shown in Figure~\ref{fig:eg4:lemmas}(A) create synthetic
clauses 25 and 28 to encode these arguments with activating literals
$\obar{v}_{11}$ and $\obar{v}_{12}$, respectively.  Clauses 31--34
then provide a proof of the lemma, stating that any assignment
$\assign$ that activates these clauses will  assign $1$ to $s_7$.
Clauses 35 and 36 state that an assignment with $\assign(x_1) = 0$
will cause the first synthetic clause to activate due to input clause
3, and it will cause the second synthetic clause to activate due to
input clause 4.  From this, clause 37 can use the lemma to state that
assigning $0$ to $x_1$ will cause $s_7$ to evaluate to $1$.  Similarly,
clauses 39 and 40 serve to activate the synthetic clauses when
$\assign(x_1) = 1$, due to input clauses 1 and 2, and clause 41 then
uses the lemma to state that assigning $1$ to $x_1$ will cause $s_7$ to
evaluate to $1$.

In this example, adding the lemma increases the proof length, but that
is only because it is such a simple formula.
